---
title: "Gun Violence - Load & Clean data"
author: "Justin Williams"
date: '2022-06-12'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-packages}
if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org" )
if(!require(janitor)) install.packages("janitor", repos = "http://cran.us.r-project.org")
if(!require(lubridate)) install.packages("lubridate", repos = "http://cran.us.r-project.org")
if(!require(remotes)) install.packages("remotes", repos = "http://cran.us.r-project.org")
remotes::install_github("Displayr/flipTime")
if(!require(dotenv)) install.packages("dotenv")
if(!require(tidycensus)) install.packages("tidycensus", repos = "http://cran.us.r-project.org")
if(!require(sf)) install.packages("sf", repos = "http://cran.us.r-project.org")
if(!require(mapview)) install.packages("mapview", repos = "http://cran.us.r-project.org")
if(!require(ggmap)) install.packages("ggmap", repos = "http://cran.us.r-project.org")
if(!require(tmaptools)) install.packages("tmaptools", repos = "http://cran.us.r-project.org")
```


## Load data

```{r load-data}
all_gun_violence <- read.csv("./data/export-ee043e5f-d7d6-4847-9aa3-7a86fcec4f5a.csv")

# clean column titles
all_gun_violence <-  all_gun_violence %>% 
  clean_names()

head(all_gun_violence)

View(all_gun_violence)
```

Load shapefile of US counties obtained from
[data.gov](https://catalog.data.gov/dataset/tiger-line-shapefile-2016-nation-u-s-current-county-and-equivalent-national-shapefile).

Can do this programmatically with **tidycensus** package and also grab some demographic statistics. But for now use this. 

```{r load-shapefile}
unzip("./data/shp/tl_2016_us_county.zip",
      exdir = "./data/shp/")

(us_counties <- st_read("./data/shp/tl_2016_us_county.shp"))
```

## Clean data

- convert `Incident Date` column to datetime
- create separate year/month/day columns
- get FIPS code for state and county
- obtain US County shapefile
- could also reverse geocode from address to get specific lat/lon, not sure that is necessary"
- actually will need to geocode from addresses:
  - first have to obtain list of state abbreviations to look up downloaded from:
  [US State Abbreviations](https://worldpopulationreview.com/states/state-abbreviations) 
  - second concatenate address, city/county, state.
  - then use **ggmap** to look up lat/lon
  - utilize script that stops at 2500, then restarts day later
  
### Convert date, year month

Convert incident date to datetime and separate columns for:

  - year
  - month
  - day
  
This will enable more specific data analysis based on date.

```{r date-column}
(all_gun_violence <- all_gun_violence %>% 
  mutate(incident_date =flipTime::AsDate(incident_date),
         year = year(incident_date),
         month = month(incident_date),
         day = day(incident_date)))
```
  
### State abbreviations

Load state abbreviations

```{r state-abbrev}
(us_state_abbr <- read_csv("./data/abbr-name-list.csv",
                          show_col_types = F))
```


### Join

Join `us_state_abbr` to `all_gun_violence` df.

```{r}
all_gun_violence_abbr <- all_gun_violence %>% 
  left_join(us_state_abbr,
            by = c("state" = "name"))
```

  
### Create address column suitable for geocoding

```{r concatenate-address-column}
(all_gun_violence_full <- all_gun_violence_abbr %>% 
  mutate(
    address_edit = str_remove(address, " block of"),
    address_edit = str_remove(address_edit, "Block of"),
    address_edit = gsub(".*(\\sand{1}\\s)","",address_edit),
    address_full = paste(address_edit, city_or_county, abbreviation,
                              sep = ", "),
    count_state = paste(city_or_county, state, sep = ", ")))
```

## Geocode addresses

Using **tmaptools** which relies on the Nominatim API pulling from OSM.

```{r geocoding-tmaptools}
test_addresses <- all_gun_violence_full$county_state[1:20]
geocode_OSM(test_addresses,
            keep.unfound = T,
            details = T,
            as.sf = T,
            geometry = "point")
```

